---
title: "Data612-Project1"
author: "Vijaya Cherukuri, Shovan Biswas, Forhad Akbar"
date: "6/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data612 - Project 1

In this first assignment, we’ll attempt to predict ratings with very little information. We’ll first look at
just raw averages across all (training dataset) users. We’ll then account for “bias” by normalizing
across users and across items.

### Our system recommends the data science books to readers. We will create a small toy dataset and use it in this recommender system

Create a dummy dataset (Toy dataset)

```{r}
row_name <- c("Vijay","Shovan","Forhad","Ryan","Steve")
col_name<-c("NLP with Python","Think Python","R for Datascience","Applied Predective Modeling", "DataScience From Scratch")
m <- matrix(c(4,3,NA,5,4,3,1,2,2,3,NA,2,NA,4,3,3,NA,3,4,NA,2,4,1,NA,4),nrow=5,byrow=TRUE, dimnames = list(row_name,col_name))
m
```

Display the matrix dimensions

```{r}
dim(m)
```

Split training and testing matrix, and replace the value from test set into "NA"
```{r}
training <- c(4,3,NA,NA,4,NA,1,2,NA,3,NA,2,NA,4,3,3,NA,NA,4,NA,2,4,1,NA,NA)
training_m <- matrix(c(4,3,NA,NA,4,NA,1,2,NA,3,NA,2,NA,4,3,3,NA,NA,4,NA,2,4,1,NA,NA),nrow=5,byrow=TRUE, dimnames = list(row_name,col_name))
testing_m <-c(5,3,2,3,4)
```

Raw_Average
```{r}
raw_average <- mean(training,na.rm=TRUE)
raw_average
```

RMSE for Training
```{r}
rmse_training <- sqrt(mean((training-raw_average)^2,na.rm = TRUE))
rmse_training
```

RMSE for Testing
```{r}
rmse_testing <- sqrt(mean((testing_m-raw_average)^2))
rmse_testing
```

Finding Bias
```{r}
training_m
user_mean <- rowMeans(training_m,na.rm=TRUE)
user_mean
```

```{r}
user_bias <- user_mean-raw_average
user_bias
```

```{r}
book_mean <- colMeans(training_m,na.rm=TRUE)
book_mean
```

```{r}
book_bias <- book_mean - raw_average
book_bias
```

Baseline Predictor
```{r}
user_bias_matrix<-matrix(c(user_bias[1],user_bias[1],user_bias[1],user_bias[1],user_bias[1],user_bias[2],user_bias[2],user_bias[2],user_bias[2],user_bias[2],user_bias[3],user_bias[3],user_bias[3],user_bias[3],user_bias[3],user_bias[4],user_bias[4],user_bias[4],user_bias[4],user_bias[4],user_bias[5],user_bias[5],user_bias[5],user_bias[5],user_bias[5]),nrow=5,byrow = TRUE)

user_bias_matrix
```

```{r}
book_bias_matrix <-matrix(c(book_bias[1],book_bias[1],book_bias[1],book_bias[1],book_bias[1],book_bias[2],book_bias[2],book_bias[2],book_bias[2],book_bias[2],book_bias[3],book_bias[3],book_bias[3],book_bias[3],book_bias[3],book_bias[4],book_bias[4],book_bias[4],book_bias[4],book_bias[4],book_bias[5],book_bias[5],book_bias[5],book_bias[5],book_bias[5]),nrow=5)

book_bias_matrix
```

RMSE for test and train
```{r}
m[is.na(m)]<-0
baseline_predictor <- m+user_bias_matrix+book_bias_matrix
baseline_predictor
```

```{r}
baseline_predictor[baseline_predictor<0] <-0
baseline_predictor[baseline_predictor>5]<- 5
baseline_predictor
```

```{r}
average_bp <- mean(baseline_predictor)
rmse_bp <- sqrt(mean((baseline_predictor-average_bp)^2))
rmse_bp
```

```{r}
rmse_bp_test <- sqrt((0+(3-2.2857)^2+(2-2.2857)^2+(3-2.2857)^2+(4-3.952581)^2)/5)
rmse_bp_test
```

Percentage Improvement for Test
```{r}
percent_improvement_test <- (1-rmse_bp_test/rmse_testing)*100
percent_improvement_test
```


Percentage Improvement for Training
```{r}
percent_improvement_training <- (1-rmse_bp/rmse_training)*100
percent_improvement_training
```

